---
title: 'POLS/CSSS 503: Problem Set 1 Solutions'
author: "Jeffrey B. Arnold"
date: "April 16, 2015"
---
```{r echo=FALSE,results='hide',message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      fig.height = 3, fig.width = 5)
```

Source the libraries that will be used

```{r libraries, message = FALSE}
library("ggplot2")
library("dplyr")
library("readr")
library("knitr")
theme_local <- theme_minimal
# nicer default binwidths
sturges_binwidth <- function(x) {
  diff(range(x)) / nclass.Sturges(x)
}
```

## Problem 1

### part a

I read in democracy using the **readr** function `read_csv`. I need to use the argument `na="."` because missing values are recorded as "."

```{r democracy,eval=FALSE}
democracy <- read.csv(file = "http://pols503.github.io/pols_503_sp15/data/democracy.csv", na = ".", stringsAsFactors = FALSE)
```

```{r democracy2,results='hide',echo=FALSE}
democracy <- read.csv(file = "democracy.csv", na = ".", stringsAsFactors = FALSE)
```

### part b

Summarize the dataset using summary
```{r summary}
summary(democracy)
```

### part c

Use `select` to remove the identifier variables, and then compute the correlation of the remaining variables with the `cor` function. The argument `use` handles how missing values in `cor` are treated. The options `"complete.obs"` will drop any case (row) with a missing value, while `"pairwise.complete.obs"` will drop pairs will missing values.
```{r cor}
democracy %>%
  select(- COUNTRY, - CTYNAME, - REGION, - YEAR) %>%
  cor(use = "pairwise.complete.obs")
```

### part d

To plot a histogram with each value of `POLLIB` in its own bin, either adjust the `binwidth` parameter of `geom_histogram`
```{r hist_pollib1}
ggplot(democracy, aes(x = POLLIB)) +
  geom_histogram(binwidth = 1) +
  xlab("Index of Political Liberty") +
  theme_local()
```
or turn `POLLIB` into a factor
```{r hist_pollib2}
ggplot(democracy, aes(x = factor(POLLIB))) +
  geom_bar() +
  xlab("Index of Political Liberty") +
  theme_local()
```
They produce slightly different looking plots, but both put each unique value of `POLLIB` (1-9) into separate plots.

### part e

A histogram of GDP per capita is

```{r hist_gdpw}
ggplot(democracy, aes(x = GDPW)) +
  geom_histogram(binwidth = sturges_binwidth(democracy$GDPW)) +
  xlab("GDP per capita") +
  theme_local()
```

### part f

A histogram of *log* GDP per capita is
```{r hist_gdpw_log}
ggplot(democracy, aes(x = log(GDPW))) +
geom_histogram(binwidth = sturges_binwidth(log(democracy$GDPW))) +
xlab("log GDP per capita") +
theme_local()
```

### part g

Create a scatterplot of political liberties against GDP per capita
```{r point_pollib_gdpw}
ggplot(democracy, aes(x = GDPW, y = POLLIB)) +
  geom_point() +
  ylab("Per capita GDP") +
  xlab("Index of Political Liberty") +
  theme_local()
```

### part h

Create the same scatterplot while jittering the points
```{r point_pollib_gdpw_jitter}
ggplot(democracy, aes(x = GDPW, y = POLLIB)) +
  geom_jitter(height = 0.5) +
  scale_y_continuous("Per capita GDP") +
  scale_x_continuous("Index of Political Liberty") +
  theme_local()
```

### part i

Create a scatterplot of political liberties against *log* GDP per capita
```{r point_pollib_log_gdpw}
ggplot(democracy, aes(x = log(GDPW), y = POLLIB)) +
  geom_jitter(height = 0.5) +
  scale_y_continuous("Log(Per capita GDP)") +
  scale_x_continuous("Index of Political Liberty") +
  theme_local()
```

### part j

A boxplot of GDP per capita for oil producing and non-oil producing nations is
```{r}
ggplot(democracy, aes(x = factor(OIL), y = GDPW)) +
  geom_boxplot() +
  scale_x_discrete("Oil Exporter") +
  theme_local()

```

### part k 

The average GDP per captical of countries at least 40% Catholic countries was 1.3 times higher than those which were less than 40% Catholic.
```{r gdpw_by_cath}
catholic_gdpw <- filter(democracy, CATH > 40)$GDPW %>% mean(na.rm = TRUE)
catholic_gdpw
all_gdpw <- mean(democracy$GDPW, na.rm = FALSE)
all_gdpw
catholic_gdpw / all_gdpw
```

### part l

The GDP per capita in countries with greater than 60% ethnolinguistic fractionalization, less than 60%, and missing ethnolinguistic fractionalization is shown in the following table:
```{r gdpw_by_elf60}
elf_summary <- democracy %>%
  mutate(high_elf60 = ELF60 > 0.6) %>%
  group_by(high_elf60) %>%
  summarise(gdpw_mean = mean(GDPW))
kable(elf_summary)
```

### part m

The median years of education for all countries in 1985 was 5.6 years.
```{r median_edt_1985}
democracy %>%
  filter(YEAR == 1985) %>%
  summarise(EDT = median(EDT, na.rm = TRUE)) %>%
  select(EDT)

```

### part n

Venezuela was the country closest (in this case, equal) to the median of the average years of education in 1985.
```{r cty_edt_median_1985}
democracy %>%
  filter(YEAR == 1985, ! is.na(EDT)) %>%
  mutate(med_edt_diff = abs(EDT - median(EDT))) %>%
  filter(med_edt_diff == min(med_edt_diff)) %>%
  select(CTYNAME, EDT)

```

### part o

The median of the average years of education for democracies in 1985 was 7.8 years.
```{r median_edt_reg_1985}
edt_reg_1985 <-
  democracy %>%
  filter(! is.na(EDT),
		 ! is.na(REG),
		 as.logical(REG),
		 YEAR == 1985)

edt_reg_1985$EDT %>% median()
```

### part p

Mauritius was democracy that were closest to the median of the average years of education for all democracies in 1985
```{r edt_reg_1985_median_country}
edt_reg_1985 %>%
  mutate(med_edt_diff = abs(EDT - median(EDT))) %>%
  filter(med_edt_diff == min(med_edt_diff)) %>%
  select(CTYNAME, EDT)
```

### part q

The 25th and 75th percentiles of the ethnolinguist fractionalization (`ELF60`) for new  and old countries (`NEWC`) is shown in the table below:

```{r}
dem_elf_tabl <-
democracy %>%
  filter(! is.na(ELF60)) %>%
  mutate(`Country Type` = ifelse(as.logical(NEWC), "new", "old")) %>%
  group_by(`Country Type`) %>%
  summarise(elf60_p25 = quantile(ELF60, probs = 0.25),
			elf60_p75 = quantile(ELF60, probs = 0.75))
kable(dem_elf_tabl)
```


## Problem 2

Answers will vary. Obviously.

