---
title: "POLS/CSSS 503: Homework 2 Solutions"
author: "Jeffrey B. Arnold"
date: "April 24, 2015"
output: html_document
---
```{r echo=FALSE,results='hide',message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      fig.height = 3, fig.width = 5)
```


Source the libraries that will be used

```{r libraries, message = FALSE}
library("ggplot2")
library("dplyr")
library("readr")
library("httr")
library("tidyr")
library("broom")
```


## Problem 1

I'll use the **httr** file to download the `csv` file directly.
```{r}
sprinters_csv <- content(GET("https://raw.githubusercontent.com/POLS503/pols_503_sp15/master/data/sprinters.csv"))
sprinters <- read_csv(sprinters_csv)
```

**a.** Create a matrix from the sprinters data using `as.matrix`, selecting only the `year` and `women` columns.
Then append a column of 1s using `cbind`.

```{r}
X <- cbind(1, as.matrix(select(sprinters, year, women)))
head(X)
```
Now, calculate $(X' X)^{-1} X' y$.
```{r}
solve(t(X) %*% X) %*% t(X) %*% sprinters$finish
```

**b.** Run a regression of finish time on year and whether the race was women.
```{r}
lm_finish <- lm(finish ~ year + women, data = sprinters)
lm_finish
```

**c.** One way to do this re-using the fitted values is with the **broom** package function `augment`, which returns the original data, augmented with fitted values and other quantities calculated from the regression.
```{r}
x_axis_scale <- scale_x_continuous("Year")
y_axis_scale <- scale_y_continuous("100m Olympic Winning Times")

sprinters <- sprinters %>%
  mutate(sex = plyr::mapvalues(women, c(0, 1), c("women", "men")))

ggplot(augment(lm_finish) %>%
       mutate(sex = plyr::mapvalues(women, c(0, 1), c("women", "men"))),
       aes(x = year, group = sex, 
           colour = sex)) +
  geom_ribbon(alpha = 0.2, colour = NA,
              mapping = aes(ymin = .fitted - 2 * .se.fit,
                            ymax = .fitted + 2 * .se.fit)) +
  geom_point(mapping = aes(y = finish)) +
  geom_line(mapping = aes(y = .fitted)) + 
  x_axis_scale + y_axis_scale + 
  theme_minimal()
```

This is a more manual way of creating the plot
```{r}
ggplot(sprinters %>% mutate(fitted = fitted(lm_finish)),
       aes(x = year, group = sex, 
           colour = sex)) +
  geom_point(mapping = aes(y = finish)) +
  geom_line(mapping = aes(y = fitted)) + 
  x_axis_scale + y_axis_scale + 
  theme_minimal()
```


**d.** Run a regression of finish on year, women, and an interaction between year and women. This is almost equivalent to running separate regressions of finish time on year for men and women.
```{r}
lm_finish_interact <- lm(finish ~ year * women, data = sprinters)
```

**e.** Redo the plot with a new fit, one for each level of women. This can easily be done with `geom_smooth`.
```{r}

ggplot(sprinters, aes(x = year, y = finish, colour = sex)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```
This can also be done using the **broom** package `tidy`.
```{r}
ggplot(augment(lm_finish_interact) %>%
       mutate(sex = plyr::mapvalues(women, c(0, 1), c("women", "men"))),
       aes(x = year, group = sex, 
           colour = sex)) +
  geom_ribbon(alpha = 0.2, colour = NA,
              mapping = aes(ymin = .fitted - 2 * .se.fit,
                            ymax = .fitted + 2 * .se.fit)) +
  geom_point(mapping = aes(y = finish)) +
  geom_line(mapping = aes(y = .fitted)) + 
  x_axis_scale + y_axis_scale + 
  theme_minimal()
```

**f.** Calculate the predicted winning times of the 100 m sprints in 2001.
```{r}
sprinters_2001 <- data_frame(year = rep(2001, 2),
                             women = c(0, 1))
predict(lm_finish_interact, newdata = sprinters_2001) 
```
Although there was no Olympics in 2001, there was a World Championship, which Maurice Greene won in 9.82 and Zhanna Pintusevich won in 10.82.

**g.** Calculate the predicted winning times of the 100 m sprints in 2001.
```{r}
sprinters_2156 <- data_frame(year = rep(2156, 2),
                             women = c(0, 1))
predict(lm_finish_interact, newdata = sprinters_2156, 
       interval = "prediction") 
```
The point predictions of the women's winning time is less than that of the men's winning time. 

**h.** No. This is far outside the range of data used to estimate the regression (`r min(sprinters$year)` to `r max(sprinters$year)`).
It is also logically incosistent, as in 3000, it would predict the winning time to be negative.
```{r}
sprinters_3000 <- data_frame(year = rep(3000, 2),
                             women = c(0, 1))
predict(lm_finish_interact, newdata = sprinters_3000, 
       interval = "prediction") 
```

## Problem 2

Let's create this data:

```{r}
data("anscombe")
anscombe2 <- anscombe %>%
  mutate(obs = row_number()) %>%
	gather(variable_dataset, value, - obs) %>%
	separate(variable_dataset, c("variable", "dataset"), sep = 1L) %>%
	spread(variable, value) %>%
	arrange(dataset, obs)
```

**a.** All four datasets have similar summary statistics for x, y and the linear relationship between x and y.

The mean  and standard deviation of $x$ 
```{r}
anscombe2 %>%
  group_by(dataset) %>%
  summarise(mean_x = mean(x),
            mean_y = mean(y),
            sd_x = sd(x),
            sd_y = sd(y),
            cor_xy = cor(x, y))
```

The regressions could be run separately
```{r}
lm(y ~ x, data = filter(anscombe2, dataset == 1))
```
but, that isn't fun.

Instead, let's take full advantage of dplyr, tidy, and broom.
Here are a couple of ways of running this code:
```{r}
anscombe2 %>%
  group_by(dataset) %>%
  do(mod = lm(y ~ x, data = .)) %>%
  mutate(intercept = coef(mod)[1],
         slope = coef(mod)[2])
```

```{r}
anscombe2 %>%
  group_by(dataset) %>%
  do(select(tidy(lm(y ~ x, data = .)), term, estimate)) %>%
  spread(term, estimate)
```


**b** Plot each dataset. They look very different, even though their coefficients are similar
```{r}
ggplot(anscombe2, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ dataset, nrow = 2) +
  theme_minimal()
  
```


## Problem 3

There will be many different responses.


* * * 

This work is licensed under a <a rel="license" href="http://creativecommons.org/licenses/by-nc-sa/4.0/">Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License</a>.
The R code is licensed under a <a rel="license" href="http://www.r-project.org/Licenses/BSD_2_clause ">BSD 2-clause</a> license.
